---
title: "The Problem with Clade-specific Sampling Fractions"
author: "Jeremy M. Beaulieu"
output: 
  pdf_document:
    fig_caption: yes
vignette: >
   %\VignetteEngine{knitr::rmarkdown}
   %\VignetteIndexEntry{The Problem with Clade-specific Sampling Fractions}
   \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
In an SSE model the sampling fraction, $f_i$ sets the initial conditions for both $D_{N_i}(0)$ -- the probability that a lineage in state $i$ at time $t$ would evolve into the extant clade N as observed -- and $1-f_i$ sets the initial conditions for $E_i(0)$ -- the probability that a lineage in state i at time $t$ would go completely extinct by the present. The important variable to pay attention to is $E_i(t)$, which does not depend on the tree structure, only on time. Thus, unlike with $D_{N_i}(t)$, at a node, N, the probability for the left and right descendants (L and R, respectively) are not combined. Instead, because the $E_i(t)$ at any node are anchored by the states, the $E_i(t)$ for the L and R descendants converges to the exact probabilities, and so it is arbitrary which set is carried down the subtending branch. However, as Moore et al (2016) pointed out with BAMM, when the diversification parameters vary between the L and R descendants, due to a shift in diversification along one branch, $E_i(t)$ does not converge at the node. 

The same problem occurs when the sampling fraction varies across the tree because it actually transforms the speciation and extinction rates to account for the missing diversity. Specifically, from Stadler (2013) the relationship between sampling fraction and speciation/extinction rates is as follows:

$$\lambda_i^* = \lambda_i / f_i; \mu_i^* = \mu_i - \lambda_i(1-1/f_i)$$
In other words, rescaling speciation and extinction in different parts of the tree due to different sampling fractions effectively creates “rate shifts” in the tree. This means that wherever in tree any two clades that differ in their sampling fractions coalesce it is not possible to simply arbitrarily choose one set of extinction probabilities to carry down in the likelihood calculation. 

```{r, eval=TRUE, echo=FALSE}
suppressPackageStartupMessages(library(deSolve))
dyn.load("bisse-ext-derivs.so")
load("cladeSamplingVariables.Rsave")
```

We can demonstrate this behavior using a straightforward example. Here we will use a three-taxon tree, with a total height of 20 time units. For ease of interpretation, we will also assume that the character states of the unsampled species are unknown completely, which means $f_i$ simply reflects the number of extant species sampled. We used the following function to conduct our branch calculations: 

```{r, eval=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
GetProbs <- function(yini, times){
        times=times
        prob.subtree.cal.full <- lsoda(yini, times, func = "maddison_DE_bisse", padded.pars, initfunc="initmod_bisse", dll = "bisse-ext-derivs", rtol=1e-8, atol=1e-8)
        probs.out <- prob.subtree.cal.full[-1,-1]
    return(probs.out)
}
```

In the case of a global sampling fraction of 50%, we can calculate the probability of the left lineage, L, which is a single branch of length 20, the extinction probabilities would be:

```{r, eval=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
times=c(0, 20)
yini <-c(E0=1-0.5, E1=1-0.5, D0=0.5, D1=0)
left.branch.probs <- GetProbs(yini, times)
left.branch.probs
```

The calculation of the extinction probabilities for the right clade, which contains two taxa that coalesce at 10 time units, is a bit more involved. We first have to combine the probabilities of the two tip branches at node R:

```{r, eval=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
times=c(0, 10)
yini <-c(E0=1-0.5, E1=1-0.5, D0=0.5, D1=0)
right.subA.probs <- GetProbs(yini, times)
right.subB.probs <- GetProbs(yini, times)

nodeR <- c(right.subA.probs[3:4] *  right.subB.probs[3:4] * c(0.1, 0.2))
#Arbritarily using the extinction probabilities from side A:
phi_A <- right.subA.probs[1:2]
```

But from here we simply use these probabilities as initial conditions when calculating probabilities for the subtending branch representing the remaining 10 time units:

```{r, eval=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
times=c(10, 20)
yini <-c(E0=phi_A[1], E1=phi_A[2], D0=nodeR[1], D1=nodeR[2])
right.branch.probs <- GetProbs(yini, times)
right.branch.probs
```

As expected, the probabilities of the left and right lineages converge to the same extinction probabilities (note, there may be slight precision issues due to our reliance on ode integration):

```{r, eval=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
round(left.branch.probs[1],4) == round(right.branch.probs[1],4)
```

Let's assume that the right clade is actually sampled rather at 25%, and the left clade is still sampled at 50%:

```{r, eval=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
times=c(0, 10)
yini <-c(E0=1-0.25, E1=1-0.25, D0=0.25, D1=0)
right.subA.probs <- GetProbs(yini, times)
right.subB.probs <- GetProbs(yini, times)
nodeR <- c(right.subA.probs[3:4] *  right.subB.probs[3:4] * c(0.1, 0.2))
#Again, arbritarily using the extinction probabilities from side A:
phi_A <- right.subA.probs[1:2]

times=c(10, 20)
yini <-c(E0=phi_A[1], E1=phi_A[2], D0=nodeR[1], D1=nodeR[2])
right.branch.probs <- GetProbs(yini, times)
right.branch.probs
```

Now, these extinction probabilities no longer converge at the root: 

```{r, eval=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff=60)}
round(left.branch.probs[1],4) == round(right.branch.probs[1],4)
```

In these situations, the likelihood is considered invalid. So, it is for these reasons that it is best to avoid using clade-specific sampling fractions and why we have removed this capability within all HiSSE functions. We recommend using either a global sampling fraction as shown above, or a global fraction for each of the states included in a given model. 

## References
Moore, B.R., S. Hohna, M.R. May, B. Rannala, and J.P. Huelsenbeck. 2016. Critically evaluating the theory and performance of Bayesian analysis of macroevolutionary mixtures. Proceedings of the National Academy of Sciences, USA, 113:9569-9574.

Stadler, T. 2013. How can we improve accuracy of macroevolutionary rate estimates? Systematic Biology, 68:321-329.
